AFQ-Insight

### 2019-12-12 reports/manuscript/abstract.tex
Incorporate Jason's round 2 edits

The white matter contains long-range connections between different
brain regions. Tractometry uses diffusion-weighted magnetic resonance
imaging (dMRI) data to quantify tissue properties (e.g. fractional
anisotropy (FA), mean diffusivity (MD), etc.), along the trajectories
of these connections\cite{yeatman2012tract}. Statistical inference from
tractometry data usually either (a) averages these quantities along
the length of the tract in each individual, or (b) performs analysis
point-by-point along each tract, with group comparisons or regression
models computed separately for each point along every one of the tracts.
In the present work, we developed a method based on the sparse group
lasso (SGL) \cite{simon2013sparse} that takes into account tissue
properties measured along all of the tracts, and selects informative
features by enforcing sparsity, not only at the level of individual
tracts and tissue properties, but also across the entire set of tracts
and all of the measured tissue properties. The sparsity penalties for
each of these constraints is identified using a nested cross-validation
scheme that guards against over-fitting and simultaneously identifies
the correct level of sparsity. We demonstrate the accuracy of the method
in two settings: in a regression setting, dMRI is used to accurately
predict ``brain age'' \cite{yeatman2014lifespan, Brown2012-so}. In this
case, the weights are distributed throughout the white matter indicating
that many different regions of the white matter change with development
and contribute to the prediction of age. In a classification setting,
patients with amyotrophic lateral sclerosis (ALS) are accurately
distinguished from matched controls\cite{sarica2017corticospinal} and
SGL automatically identifies FA in the corticospinal tract as important
for this classification -- correctly finding the parts of the white
matter known to be affected by the disease. Thus, SGL makes it possible
to leverage the multivariate relationship between diffusion properties
measured along multiple tracts to make accurate predictions of subject
characteristics while simultaneously discovering the most relevant
features of the white matter for the characteristic of interest.

>>>>

The white matter contains long-range connections between different
brain regions and the organization of these connections hold important
implications for brain function in health and disease. Tractometry
uses diffusion-weighted magnetic resonance imaging (dMRI) data
to quantify tissue properties (e.g. fractional anisotropy (FA),
mean diffusivity (MD), etc.), along the trajectories of these
connections\cite{yeatman2012tract}. Statistical inference from
tractometry data usually either (a) averages these quantities along
the length of the tract in each individual, or (b) performs analysis
point-by-point along each tract, with group comparisons or regression
models computed separately for each point along every one of the tracts.
In the present work, we developed a method based on the sparse group
lasso (SGL) \cite{simon2013sparse} that takes into account tissue
properties measured along all of the tracts, and selects informative
features by enforcing sparsity, not only at the level of individual
tracts, but also across the entire set of tracts and all of the measured
tissue properties. The sparsity penalties for each of these constraints
is identified using a nested cross-validation scheme that guards
against over-fitting and simultaneously identifies the correct level of
sparsity. We demonstrate the accuracy of the method in two settings:
in a regression setting, dMRI is used to accurately predict ``brain
age'' \cite{yeatman2014lifespan, Brown2012-so}. In this case, the
weights are distributed throughout the white matter indicating that
many different regions of the white matter change with development
and contribute to the prediction of age. In a classification setting,
patients with amyotrophic lateral sclerosis (ALS) are accurately
distinguished from matched controls\cite{sarica2017corticospinal} and
SGL automatically identifies FA in the corticospinal tract as important
for this classification -- correctly finding the parts of the white
matter known to be affected by the disease. Thus, SGL makes it possible
to leverage the multivariate relationship between diffusion properties
measured along multiple tracts to make accurate predictions of subject
characteristics while simultaneously discovering the most relevant
features of the white matter for the characteristic of interest.

### 2019-12-12 reports/manuscript/sections/conclusion.tex
Incorporate Jason's round 2 edits

Specifically, we demonstrated that the algorithm correctly detects the fact that
ALS, which is a disease of lower motor neurons, is localized to the
cortico-spinal tract. This recapitulates the results of previous analysis of
these same data, using a targeted ROI-based approach
\cite{sarica2017corticospinal}. In contrast, in analysis of biological age, the
coefficients identified by the algorithm are very widely distributed across many
parts of the white matter, mirroring previous results with this dataset that
show a large and continuous distribution of life-span changes in white matter
properties \cite{yeatman2014lifespan}.

@@

SGL has many other potential applications in neuroscience, because of the
hierarchical and grouped nature of many of the data -- that are collected in
multiple sample points within anatomically-defined areas. For example, this
method may be useful to understand the relationship between fMRI recordings and
behavior, where activity in each voxel may co-vary with voxels within the same
anatomical region and form features and groups of features. Similarly,
large-scale multi-electrode recordings of neural activity in awake behaving
animals are becoming increasingly feasible \cite{steinmetz2018distributed,
Jun2017-gv} and these recordings can form features (neurons) and groups (neurons
within an anatomical region). More ambitiously perhaps, this approach may be
used to understand the role of correlations in so-called resting-state fMRI
time-series and behavior, where pairwise correlations between voxels in
different anatomical regions are features in the matrix and features may be
grouped by pairs of anatomical regions. Given the large number of voxels in the
surface of the gray matter, and given that correlations increase the number of
features by a factor of $n^2$ this would be a challenging problem to solve using
SGL.

>>>>

Specifically, we demonstrated that the algorithm correctly detects
the fact that ALS, which is a disease of lower motor neurons, is
localized to the cortico-spinal tract. This recapitulates the results
of previous analysis of these same data, using a targeted ROI-based
approach \cite{sarica2017corticospinal}. In contrast, for the analysis
of biological age, the coefficients identified by the algorithm are very
widely distributed across many parts of the white matter, mirroring
previous results with this dataset (and others) that show a large and
continuous distribution of life-span changes in white matter properties
\cite{yeatman2014lifespan}.

@@

SGL has many other potential applications in neuroscience, because
of the hierarchical and grouped nature of many data types that are
collected in multiple sample points within anatomically-defined areas.
For example, this method may be useful to understand the relationship
between fMRI recordings and behavior, where activity in each voxel
may co-vary with voxels within the same anatomical region and form
features and groups of features. Similarly, large-scale multi-electrode
recordings of neural activity in awake behaving animals are becoming
increasingly feasible \cite{steinmetz2018distributed, Jun2017-gv} and
these recordings can form features (neurons) and groups (neurons within
an anatomical region). More ambitiously perhaps, this approach may be
used to understand the role of correlations in so-called resting-state
fMRI time-series and behavior, where pairwise correlations between
voxels in different anatomical regions are features in the matrix and
features may be grouped by pairs of anatomical regions. Given the large
number of voxels in the surface of the gray matter and given that
correlations increase the number of features by a factor of $n^2$, this
would pose a challenging problem to solve using SGL.

### 2019-12-12 reports/manuscript/sections/introduction.tex
Incorporate Jason's round 2 edits

Using this fact, one of the most powerful methods currently available to put
macro- and micro-structure together is \emph{tractometry} (also known as
tract-based morphometry): assessment of the physical properties of the white
matter along specific tracts \cite{Bells2011-cf}. Though there are several
different available implementations of this overall idea, the principles are
similar \cite{yeatman2012tract, Yendiki2011-ay, Wassermann2016-iv,
ODonnell2009-uu}: tractometry begins by delineating the parts of the white
matter that belong to different major tracts or axonal nerve fiber bundles, such
as the corticospinal tract or arcuate fasciculus, and samples the biophysical
properties (such as fractional anisotropy or mean diffusivity) along the length
of these bundles. In some cases, tissue properties along the length of each
bundle are summarized by taking the mean along each bundle, but there is a large
body of evidence showing that there is systematic variability within each
bundle. This justifies retaining the individual samples along the length of each
bundle \cite{yeatman2012tract, colby2012, ODonnell2009-uu}. While this retains
important information about each individual's white matter, it also presents
statistical challenges due to the dimensionality of the data. Based on
tractometry, researchers may choose to compare different individuals to each
other. This is usually done following one of the following approaches:

@@

\item Mass univariate approaches: In this approach comparisons between groups or
across individuals are done at each node of each tract, for each one of the
diffusion metrics available at that point. This approach is exhaustive, but
statistical power is compromised by a multiple comparison problem. Different
approaches can be taken to resolving this challenge. For example, Colby and
colleagues \cite{colby2012} used a non-parametric resmapling approach to correct
for family-wise error across the different possible comparisons
\cite{Nichols2002-zu, Nichols2003-yy}.

@@

Generally speaking analysis methods should balance predictive accuracy with
descriptive power \cite{Murdoch2019-ax, Breiman2001-uz}. Accordingly,
tractometry analysis should simultaneously capitalize on all the data across all
tracts to make the best possible prediction, while also retaining and surfacing
spatial information about the locations that are most informative for a
prediction. In the present work, we developed a novel framework for analysis of
tractometry that simultaneously selects the features for analysis, and fits a
model to these features. We use a linear modeling approach, which aims to
predict phenotypical variance in a group of subjects, based on a linear
combination of the features estimated with tractometry. Using this approach, we
need to first deal with the large and asymmetric dimensionality of the data:
tractometry data usually has many more features (i.e., number of measurements
per individual) than samples (number of subjects), which makes inferences from
the data about phenotypical differences between individuals ill-posed. This
regime is the target of several statistical learning techniques, and is often
solved by various forms of regularization. For example, Tikhonov regularization
shrinks the solution such that the sum of squared contributions from the
individual features are minimized \cite{Hoerl2000-ij}. Another solution to the
problem is provided by the Lasso algorithm, which instead minimizes the sum of
the absolute values of contributions of each feature \cite{Tibshirani1996-qs}.
This tends to shrink to zero the contributions of many of the features,
providing results that are both accurate and interpretable. When additional
structure is available in the organization of the data, regularization
algorithms can take advantage of this structure. For example, if the features
lend themselves to a natural division into different groups, the group lasso
(GL) can be used to select groups of features, rather than individual features
\cite{Yuan2006-ky}. The Sparse Group Lasso (SGL) elaborates on this idea by
providing control both of group sparsity, as well as overall sparsity of the
solutions \cite{simon2013sgl}. Because the features measured with tractomery
lend themselves to grouping based on the tracts from which each measurement is
taken, GL and SGL provide a useful tool linear model fitting in problems of this
form. Using linear regression as the basis for analysis is useful for both
regression tasks, in which a continuous measure (e.g., age) is modeled, as well
as for classification tasks, where a discrete property is modeled (e.g., whether
an individual has a particular disease). Moreover, use of SGL helps derive
informative properties of the human white matter, both in cases where the
underlying causes of variance are sparsely concentrated (e.g., in one particular
white matter tract) and in cases where biological substrates of differences
between individuals are more widespread. This is because SGL is controlled by
meta-parameters that set the degree to which group sparsity and overall sparsity
are applied. These metaparameters can be set empirically, based on the data,
through a process of nested cross-validation. We demonstrate these properties of
the algorithm in data from previously published studies
\cite{sarica2017corticospinal, yeatman2014lifespan}.

>>>>

Leveraging this fact, one of the most powerful methods currently
available to put macro- and micro-structure together is
\emph{tractometry}: assessment of the physical properties of the white
matter along specific tracts \cite{Bells2011-cf}. Though there are
several different available implementations of this overall idea,
the principles are similar \cite{yeatman2012tract, Yendiki2011-ay,
Wassermann2016-iv, ODonnell2009-uu}: tractometry begins by delineating
the parts of the white matter that belong to different major tracts or
axonal nerve fiber bundles, such as the corticospinal tract or arcuate
fasciculus, and samples the biophysical properties (such as fractional
anisotropy or mean diffusivity) along the length of these bundles.
In some cases, tissue properties along the length of each bundle are
summarized by taking the mean along each bundle, but there is a large
body of evidence showing that there is systematic variability along
the trajectory of each bundle. This justifies retaining the individual
samples along the length of each bundle \cite{yeatman2012tract,
colby2012, ODonnell2009-uu}. However, while this retains important
information about each individual's white matter, it also presents
statistical challenges due to the dimensionality of the data. Based on
tractometry, researchers may choose to compare different individuals
to each other. This is usually done following one of the following
approaches:

@@

\item Mass univariate approaches: In this approach comparisons between
groups or across individuals are done independently at each node
of each tract, for each one of the diffusion metrics available at
that point. This approach is exhaustive, but statistical power is
compromised by a multiple comparison problem. Different approaches
can be taken to resolving this challenge. For example, Colby and
colleagues \cite{colby2012} used a non-parametric resmapling approach to
correct for family-wise error across the different possible comparisons
\cite{Nichols2002-zu, Nichols2003-yy}.

@@

Generally speaking, analysis methods should balance predictive
accuracy with descriptive power \cite{Murdoch2019-ax, Breiman2001-uz}.
Accordingly, tractometry analysis should simultaneously capitalize on
all the data across all tracts to make the best possible prediction,
while also retaining and elucidating spatial information about the
locations that are most informative for a prediction. In the present
work, we developed a novel framework for analysis of tractometry that
simultaneously selects the features for analysis, and fits a model
to these features. We use a linear modeling approach, which aims to
predict phenotypical variance in a group of subjects, based on a linear
combination of the features estimated with tractometry.

Using this approach, we first need to deal with the large and
asymmetric dimensionality of the data: tractometry data usually has
many more features (i.e., number of measurements per individual) than
samples (number of subjects), which makes inferences from the data
about phenotypical differences between individuals ill-posed. This
regime is the target of several statistical learning techniques, and
is often solved by various forms of regularization. For example,
Tikhonov regularization shrinks the solution such that the sum of
squared contributions from the individual features are minimized
\cite{Hoerl2000-ij}. Another solution to the problem is provided by
the Lasso algorithm, which instead minimizes the sum of the absolute
values of contributions of each feature \cite{Tibshirani1996-qs}. This
tends to shrink to zero the contributions of many of the features,
providing results that are both accurate and interpretable. When
additional structure is available in the organization of the data,
regularization algorithms can take advantage of this structure. For
example, if the features lend themselves to a natural division into
different groups, the group lasso (GL) can be used to select groups
of features, rather than individual features \cite{Yuan2006-ky}.
The Sparse Group Lasso (SGL) elaborates on this idea by providing
control both of group sparsity, as well as overall sparsity of the
solutions \cite{simon2013sgl}. Because the features measured with
tractomery lend themselves to grouping based on the tracts from which
each measurement is taken, GL and SGL could provide a useful tool
for linear model fitting in problems of this form. Here we, first,
develop an implementation of SGL that is well suited to the analysis of
tractometry data and, second, demonstrate the power and flexibility of
this approach by applying it to both classification (disease diagnosis)
and continuous prediction (age) problems from previously published
studies \cite{sarica2017corticospinal, yeatman2014lifespan}.
